---
title: 'Calculating biodiversity metrics from rcoalescence simulations'
author: "Sam Thompson"
date: "10/05/2020"
output: 
  pdf_document: 
    keep_tex: yes
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{rcoalescence-estimating-biodiversity-loss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
  library(rcoalescence)
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(viridis)
})
```

## Introduction

In this vignette, we first run a small simulation and then demonstrate how some simple biodiversity metrics can be extracted using the in-built functions, or calculated in a few lines of R.

First, we run a spatial simulation on a "null" landscape (with 1 in every cell) of dimensions 1000 by 1000 (or load the simulation if it already exists).

```{r}
# Path to the output database
output_db <- file.path("output", paste0("data_", 1, "_", 1001, ".db"))
sim <- SpatialTreeSimulation$new()
sim$setSimulationParameters(
  task = 1,
  seed = 1001,
  output_directory = "output",
  min_speciation_rate = 0.0001,
  fine_map_file = "null",
  fine_map_x_size = 1000,
  fine_map_y_size = 1000,
  deme = 1,
  sigma = 2,
  landscape_type = "closed"
)
if (file.exists(output_db)) {
  sim$setOutputDatabase(output_db)
} else {
  sim$runSimulation()
  sim$applySpeciationRates(
    speciation_rates = c(
      0.0001,
      0.001,
      0.01,
      0.1
    ),
    use_spatial = TRUE
  )
  sim$output()
}
```

## Extracting the list of communities

Each speciation rate applied to our simulation generates a different community. The community reference relates communities to the associated parameters and is extracted using `sim$getCommunityReferences()`. Often, it is useful to rename community_reference to reference to join with other tables.

```{r}
community_parameters <- sim$getCommunityReferences() %>%
  rename(community_reference = reference)
```

```{r echo=FALSE}
knitr::kable(community_parameters,
  caption = "Community parameters dataframe, as extracted from the simulation using \texttt{getCommunityReferences()}.",
  booktabs = TRUE
) %>%
  kable_styling()
```

## Species abundance distributions

Using the `getSpeciesAbundances()` function and joining it to the community parameters (each speciation rate in this example), the
species abundances distributions for each speciation rate can be extracted.

```{r}
species_abundances <- sim$getSpeciesAbundances(sim$getCommunityReferences()$reference) %>%
  left_join(community_parameters,
    by = "community_reference"
  )
```

This produces the following dataframe.

```{r echo=FALSE}
knitr::kable(species_abundances %>% head(20),
  caption = "Species abundances dataframe, as extracted from the simulation (first 20).",
  booktabs = TRUE
) %>%
  kable_styling(font_size = 7)
```

The species abundances can be binned into log2 classes and then counted, so that the number of species within each bin is totalled. This provides the data necessary for \cite{Preston1960} plots.

```{r}
binned_species_abundances <- species_abundances %>%
  filter(no_individuals > 0) %>%
  mutate(log2_abundance = floor(log2(no_individuals))) %>%
  group_by(
    community_reference,
    speciation_rate,
    log2_abundance
  ) %>%
  summarise(no_species = n())
```

These species abundances distributions can now be plotted.

```{r echo=FALSE, warning=FALSE, fig.fullwidth=TRUE, fig.width=6, fig.cap="Species abundance distributions for each community as Preston plots."}
sad_plot <- binned_species_abundances %>%
  ggplot() +
  geom_line(aes(
    x = log2_abundance, y = no_species,
    colour = as.factor(speciation_rate),
    linetype = as.factor(speciation_rate)
  )) +
  scale_x_continuous(expression(paste(
    "Abundance class (",
    log[2],
    ")"
  )),
  trans = scales::log2_trans(),
  labels = scales::trans_format(
    "log2",
    scales::math_format(.x)
  )
  ) +
  scale_y_continuous("Number of species") +
  scale_colour_brewer("Speciation\nrate",
    palette = 2, type = "qual",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  scale_linetype("Speciation\nrate",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  theme_classic() +
  theme(aspect.ratio = 0.7)
sad_plot
```




## Species-area relationships

One method of calculating the species-area relationship involves first extracting the locations of each species, then counting the number of species within an increasingly large square. A few lines of code can accomplish this task. For simplicity, we are only looking at squares within the landscape which include at least one individual.

First, define some functions for calculating SARs.

```{r}
#' Calculates the richness of species within a given size in the
#' landscape, counted in cells.
#' Randomly samples from the landscape a number of times to get
#' coordinates for calculating the area.
#' Returns an array of species richness values
calculate_richness_in_area <- function(df, a, sample_number) {
  min_x <- min(df$x)
  max_x <- max(df$x)
  min_y <- min(df$y)
  max_y <- max(df$y)

  i <- 1
  rand_xs <- sample(min_x:(max_x - a),
    sample_number,
    replace = TRUE
  )
  rand_ys <- sample(min_y:(max_y - a),
    sample_number,
    replace = TRUE
  )
  species_number <- vector("list", sample_number)

  count_species <- function(i) {
    x_ <- rand_xs[i]
    y_ <- rand_ys[i]
    no_species <- df %>%
      filter(
        x >= x_,
        x < x_ + a,
        y >= y_,
        y < y_ + a,
      ) %>%
      select(species_id) %>%
      distinct() %>%
      count() %>%
      pull()
    return(no_species)
  }
  species_number <- lapply(seq(sample_number), count_species)
  return(unlist(species_number))
}

#' Calculate the min, mean, max and lower and upper 25% values
calculate_min_mean_max <- function(df, a, sample_number) {
  species_richness <- calculate_richness_in_area(df,
    a,
    sample_number = sample_number
  )
  return(data.frame(
    area = a,
    min_richness = min(species_richness),
    mean_richness = mean(species_richness),
    max_richness = max(species_richness)
  ))
}

#' Loop over a selection of possible sizes in the landscape
calculate_species_richness_all_areas <- function(df,
                                                 min_size = 10,
                                                 interval = 50,
                                                 sample_number = 100) {
  max_size <- min(max(df$x) - min(df$x), max(df$y) - min(df$y))
  output_richness <- bind_rows(lapply(seq(
    min_size,
    max_size,
    interval
  ),
  calculate_min_mean_max,
  df = df,
  sample_number = sample_number
  ))
  return(output_richness)
}
```


Now, run the calculation for each speciation rate. This function takes around 30 seconds to a minute.

```{r}
sar <- NULL
for (ref in sim$getCommunityReferences()$reference) {
  sar_tmp <- calculate_species_richness_all_areas(sim$getSpeciesLocations(ref),
    sample_number = 10
  ) %>%
    mutate(reference = ref)
  if (is.null(sar)) {
    sar <- sar_tmp
  } else {
    sar <- sar %>% bind_rows(sar_tmp)
  }
}
sar <- sar %>%
  left_join(sim$getCommunityReferences(),
    by = "reference"
  )
```

Show the results as a table.

```{r echo=FALSE}
knitr::kable(sar %>% head(20),
  caption = "Species richness as a function of area, summarised (top 20).",
  booktabs = TRUE,
  longtable = TRUE
) %>%
  kable_styling(latex_options = c("repeat_header"))
```

And plot the SAR.

```{r echo=FALSE, warning=FALSE, fig.fullwidth=TRUE, fig.width=6, fig.cap="Species-area relationships for each speciation rate."}
sar_plot <- sar %>%
  ggplot() +
  geom_ribbon(aes(
    x = area,
    ymin = min_richness,
    max = max_richness,
    fill = as.factor(speciation_rate)
  ), alpha = 0.5) +
  geom_line(aes(
    x = area,
    y = mean_richness,
    colour = as.factor(speciation_rate),
    linetype = as.factor(speciation_rate)
  )) +
  scale_x_continuous("Log10(Area)",
    trans = "log10"
  ) +
  scale_y_continuous("Log10(number of species)",
    trans = "log10",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  scale_colour_brewer("Speciation\nrate",
    palette = 2, type = "qual",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  scale_fill_brewer("Speciation\nrate",
    palette = 2, type = "qual",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  scale_linetype("Speciation\nrate",
    labels = scales::math_format(format = function(x) {
      log10(as.numeric(x))
    })
  ) +
  theme_classic() +
  theme(aspect.ratio = 0.7)
sar_plot
```
