---
title: "Micro-benchmarks for rcoalescence"
output: 
  pdf_document: 
    keep_tex: yes
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{rcoalescence-estimating-biodiversity-loss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib  
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
suppressPackageStartupMessages({
  library(rcoalescence)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(tiff)
  library(reshape2)
  library(viridis)
  library(ggpubr)
  library(kableExtra)
})
output_dir <- file.path("output", "bench")
figure_directory <- "figures"
if (!dir.exists(figure_directory)) {
  dir.create(figure_directory)
}
temp_dir <- "temp"
if (dir.exists(temp_dir)) {
  unlink(temp_dir, recursive = TRUE)
}
dir.create(temp_dir)
```

The microbenchmark code is provided at https://github.com/thompsonsed/pycoalescence_bench. The graphs below show simulation time against a number of different simulation parameters. These values are only given as indicators for smaller-scale simulations. Generally, simulation time increases when:

- speciation rate decreases
- the number of simulated individuals increases
- dispersal increases (except when dispersal is exceptionally small)
- the landscape area increases



```{r echo=FALSE, warning=FALSE, fig.fullwidth=TRUE, fig.width=6, fig.height=3, fig.cap="Mean simulation time (averaged over 4 repeats) as a function of the landscape size. Colours represent the per-capita speciation rate. Different facets indicate the relative scale of the surrounding landscape within which individuals can move, before a hard boundary is hit."}

microbench_df <- read.csv(file.path(output_dir, "microbenchmarks.csv")) %>%
  mutate(scale = as.character(scale)) %>%
  bind_rows(read.csv(file.path(output_dir, "microbenchmarks_infinite.csv")) %>%
    mutate(scale = "infinite")) %>%
  mutate(scale = factor(ifelse(scale == "infinite", "Infinite",
    paste0(scale, "x scale")
  ),
  levels = c("3x scale", "9x scale", "25x scale", "Infinite")
  ))
microbench_df %>%
  filter(sigma == 2) %>%
  group_by(
    size,
    scale,
    speciation_rate,
    sigma
  ) %>%
  summarise(
    min_time = min(time),
    max_time = max(time),
    mean_time = mean(time),
  ) %>%
  ggplot() +
  geom_ribbon(aes(
    x = size^2,
    ymin = min_time,
    ymax = max_time, fill = speciation_rate, group = speciation_rate
  ), alpha = 0.5) +
  geom_line(aes(
    x = size^2, y = mean_time,
    colour = speciation_rate,
    group = speciation_rate
  )) +
  theme_classic() +
  facet_grid(. ~ scale) +
  theme(aspect.ratio = 1.2) +
  scale_x_continuous("Size of landscape (number of individuals)",
    trans = "log2", labels = scales::math_format(2^.x, format = log2)
  ) +
  scale_y_continuous("Mean simulation time (seconds)", trans = "log2") +
  scale_colour_viridis("Speciation rate",
    trans = "log10",
    labels = scales::math_format(10^.x, format = function(x) {log10(x)}),
    option = "plasma"
  ) +
  scale_fill_viridis("Speciation rate",
                     trans = "log10",
                     option = "plasma", 
                     labels = scales::math_format(10^.x, format = function(x) {log10(x)}))
```

```{r echo=FALSE, warning=FALSE, fig.fullwidth=TRUE, fig.width=6, fig.height=3, fig.cap="Mean simulation time (averaged over 4 repeats) as a function of per-capita speciation rate. Colours represent the size of the landscape in number of individuals. Different facets indicate the relative scale of the surrounding landscape within which individuals can move, before a hard boundary is hit."}
microbench_df %>%
  filter(sigma == 2) %>%
  group_by(size, scale, speciation_rate, sigma) %>%
  summarise(
    min_time = min(time),
    max_time = max(time),
    mean_time = mean(time),
  ) %>%
  ungroup() %>%
  mutate(size = size^2) %>%
  ggplot() +
  geom_ribbon(aes(x = speciation_rate, ymin = min_time, ymax = max_time, fill = size, group = size), alpha = 0.5) +
  geom_line(aes(
    x = speciation_rate,
    y = mean_time,
    colour = size,
    # linetype=as.factor(scale),
    group = size
  )) +
  scale_x_log10("Speciation rate", 
                labels = scales::math_format(10^.x, format = function(x) {log10(x)})) +
  scale_y_continuous("Mean simulation time (seconds)", trans = "log2") +
  theme_classic() +
  theme(aspect.ratio = 1.2) +
  facet_grid(. ~ scale) +
  scale_colour_viridis("Size", 
                       option = "magma", 
                       trans = "log2", 
                       end = 0.9, 
                       labels = scales::math_format(2^.x, format = function(x) {log2(x)})) +
  scale_fill_viridis("Size", 
                     option = "magma", 
                     trans = "log2", 
                     end = 0.9, labels = scales::math_format(2^.x, format = function(x) {log2(x)}))
```

```{r echo=FALSE}
kable(microbench_df %>%
  filter(scale == "Infinite") %>%
  group_by(size, scale, speciation_rate, sigma) %>%
  summarise(
    `min time` = round(min(time), 2),
    `mean time` = round(mean(time), 2),
    `max time` = round(max(time), 2)
  ) %>%
  ungroup() %>%
  mutate(size = size^2) %>%
  rename(
    `speciation rate` = speciation_rate,
    `number of individuals` = size
  ),
caption = "Summary table of simulation times from the microbenchmark",
longtable = TRUE,
booktabs = TRUE
) %>%
  kable_styling(latex_options = c("repeat_header"))
```

```{r echo=FALSE, warning=FALSE, fig.fullwidth=TRUE, fig.width=6, fig.height=3, fig.cap="Mean simulation time (averaged over 4 repeats) as a function of dispersal rate (sigma) on infinite landscapes. Colours represent the per-capita speciation rate. Different facets indicate the number of individuals in the sampled landscape."}
microbench_df %>%
  filter(scale == "Infinite") %>%
  group_by(size, scale, speciation_rate, sigma) %>%
  summarise(
    min_time = min(time),
    max_time = max(time),
    mean_time = mean(time),
  ) %>%
  ungroup() %>%
  mutate(size = size^2) %>%
  filter(size < 5000) %>%
  ggplot() +
  geom_ribbon(aes(
    x = sigma,
    ymin = min_time,
    ymax = max_time,
    fill = speciation_rate,
    group = speciation_rate
  ), alpha = 0.5) +
  geom_line(aes(
    x = sigma,
    y = mean_time,
    colour = speciation_rate,
    group = speciation_rate
  )) +
  scale_x_continuous("Dispersal rate (sigma)", trans="log2") +
  scale_y_continuous("Mean simulation time (seconds)", trans = "log2") +
  theme_classic() +
  theme(aspect.ratio = 1.2) +
  scale_colour_viridis("Speciation rate",
                       option = "plasma", 
                       trans = "log10",
                       labels = scales::math_format(10^.x, format = function(x) {log10(x)})) +
  scale_fill_viridis("Speciation rate",
                     option = "plasma",
                     trans = "log10",
                     labels = scales::math_format(10^.x, format = function(x) {log10(x)})) +
  facet_grid(. ~ size)
```
